### 2021-04-20 introduction.tex
target ISSRE

Combining a differential approach and mutation analysis offers a novel way to compare static analysis tools, one useful to users wishing to select a good tool or set of tools, to researchers interested in the impact of precision/soundness trade-offs or different intermediate languages, and to developers of static analysis tools hoping to improve their tools by examining missed faults.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.35\columnwidth]{example.png}
  \caption{Mutants killed by three static analysis tools.}
  \label{fig:examplevenn}
\end{figure}

@@

More concretely, consider the code in Figure \ref{fig:sol424intro} \cite{solintro}.  The Universal Mutator tool \cite{universalmutator,regexpMut}, which has been extensively tuned for Solidity's grammar (though not to target any particular vulnerabilities), and is the only smart contract mutation tool referenced in the Solidity documentation (\url{https://solidity.readthedocs.io/en/v0.5.12/resources.html}), produces seven valid, non-redundant (by Trivial Compiler Equivalence \cite{TCE}) mutants for it.  Both the public version of Slither \cite{slither} and SmartCheck \cite{smartcheck} (two popular smart contract static analysis tools) produce a small number (three and two, respectively) of low-severity, informational, warnings for this code.  Both tools also detect four of the seven mutants (here the number of warnings increases, and the additional warnings are clearly driven by the mutation change) .  However, only one of the mutants detected is common to both tools: both tools detect changing the {\tt return} statement in the {\tt get} function to a call to {\tt selfdestruct} the smart contract, deleting it.  Slither, but not SmartCheck, also detects replacing the assignment of {\tt storedData} in {\tt set} with either a {\tt selfdestruct} or {\tt revert}, or simply removing it altogether.  SmartCheck, on the other hand, detects removing the {\tt return} in {\tt get} or replacing it with a {\tt revert}, or removing the {\tt public} visibility modifier for {\tt get} \footnote{Slither's ``missing return'' detector is only available in the private version of slither, or through the {\tt crytic} service provided by Trail of Bits.}.  If we restrict our analysis to findings with a severity greater than informational, SmartCheck detects no mutants of the contract, while Slither still reports that some mutants \emph{allow an arbitrary caller to cause the contract to destroy itself}.  Given that both tools, ignoring informational results, detect no problems with the original code, and only Slither detects any problems with the mutated code, we can say that Slither performs better for static analysis of mutants for this contract.

@@

A general objection to our approach is that mutants may differ substantially from ``real'' faults, in some way.  This is certainly true, in a sense \cite{GopinathMutants}, but for static analysis purposes we believe it does not matter.   The real risk is that some mutation operators align with patterns a particular tool identifies, biasing the evaluation in favor of that tool.  Such faults may be dis-proportionately present in mutants vs. real code.  However, we consider this unlikely.  The vast majority of applied mutation operations for all of our experiments were highly generic, and do not plausibly represent a pattern in which some tool might specialize.  Code deletions, the most common kind of mutation by far, leave no ``trace'' for a tool to match against, but only an omission, so cannot be subject to this concern.  Changing arithmetic and comparison operators and numeric constants (incrementing, decrementing, or changing to 0 or 1) account for most of the non-statement-deletion mutants.  It is difficult to imagine how any useful tool could unfairly identify these kinds of mutants.

@@

\begin{itemize}[labelsep=3pt,leftmargin=12pt]
\item We propose a differential approach to comparing static analysis tools based on the insight that program mutants provide an automated source of simple, easy-to-understand program changes that are likely faults.
\item We propose a definition of mutant killing for static analysis.
\item We introduce a simple scheme for prioritizing mutants that helps users understand and use the results of analysis.
\item We apply our method to an extensive, in-depth comparison of three Solidity smart contract analysis tools, and show how prioritization allowed us to easily identify (and build) three new detectors for the most effective of these tools.
\item We further provide results for comparisons of popular Java and Python static analysis tools, showing the general usefulness of our methods, and giving a new picture of the comparative effectiveness of these tools.
\end{itemize}

 While there are certainly limitations to using differential mutation analysis to compare and extend static analysis tools, the method scales to basing comparisons on large numbers of real software source files and very large numbers of faults, but still has some of the advantages of humans establishing ground truth for tool findings.

@@

\begin{enumerate}
\item Run each tool on the unmutated source code target(s), and determine the \emph{baseline}: the number of (non-informational/stylistic) findings produced.
\item Generate mutants of the source code and run each tool on each mutant.  Consider a mutant killed if the number of findings for the mutated code is greater than the number for the baseline, un-mutated code.
\item Compute, for each tool, the \emph{mutant ratio}:  the mutation score ($\frac{|\mathit{killed}|}{|\mathit{mutants}|}$) divided by (mean) baseline.  If the (mean) baseline is zero, use a baseline equal to either one or the lowest non-zero baseline for any tool in the comparison set\footnote{Unless the number of programs used for evaluation is small, a zero mean baseline for a tool is extremely unlikely to arise in practice.}.
\item (Optional): Discard all mutants not killed by at least one tool and all mutants killed by all tools.  What remains allows \emph{differential} analysis.
Examine the remaining mutants in the difference in \emph{prioritized} order to qualitatively understand tool differences, or improve a tool.
\end{enumerate}

@@

In order to apply FPF ranking to examining mutants, we implemented a
simple, somewhat \emph{ad hoc} distance metric and FPF ranker.  Our metric $d$ is
the sum of a set of measurements.  First, it adds a similarity
ratio based on Levenshtein distance \cite{lev} for (1) the \emph{changes} (Levenshtein edits) from
the original source code elements to
the two mutants,  (2) the two original source code elements changed (in
general, lines), and (3) the actual output mutant code.  These are
weighted with multipliers of 5.0, 0.1, and 0.1, respectively; the type
of change (mutation operator, roughly) dominates this part of the
distance, because it best describes ``what the mutant did''; however,
because many mutants will have the same change (e.g., changing {\tt +}
to {\tt -}, the other ratios also often matter.
Our metric also incorporates a measure of the distance in the source
code between the locations of two mutants.  If the mutants are to
different files, this adds 0.5 to the distance; it also adds 0.25
times the number of source lines separating the two mutants if they
are in the same file, divided by 10, but caps the amount added at
0.25.

>>>>

Combining a differential approach and mutation analysis offers a novel way to compare static analysis tools, one useful to users wishing to select a good tool or set of tools, to researchers interested in the impact of precision/soundness trade-offs or different intermediate languages, and to developers of static analysis tools hoping to improve their tools.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.30\columnwidth]{example.png}
  \caption{Mutants killed by three static analysis tools.}
  \label{fig:examplevenn}
\end{figure}

@@

More concretely, consider the code in Figure \ref{fig:sol424intro} \cite{solintro}.  The Universal Mutator tool \cite{universalmutator,regexpMut}, which has been extensively tuned for Solidity's grammar (though not to target any particular vulnerabilities), and is the only smart contract mutation tool referenced in the Solidity documentation (\url{https://solidity.readthedocs.io/en/v0.5.12/resources.html}), produces seven valid, non-redundant (by Trivial Compiler Equivalence \cite{TCE}) mutants for it.  Both the public version of Slither \cite{slither} and SmartCheck \cite{smartcheck} (two popular smart contract static analysis tools) produce a small number (three and two, respectively) of low-severity, informational, warnings for this code.  Both tools also detect four of the seven mutants (here the number of warnings increases, and the additional warnings are clearly driven by the mutation change) .  However, only one of the mutants detected is common to both tools: both tools detect changing the {\tt return} statement in the {\tt get} function to a call to {\tt selfdestruct} the smart contract, deleting it.  Slither, but not SmartCheck, also detects replacing the assignment of {\tt storedData} in {\tt set} with either a {\tt selfdestruct} or {\tt revert}, or simply removing it altogether.  SmartCheck, on the other hand, detects removing the {\tt return} in {\tt get} or replacing it with a {\tt revert}, or removing the {\tt public} visibility modifier for {\tt get} \footnote{Slither's ``missing return'' detector is only available in the private version of slither, or through the {\tt crytic} service provided by Trail of Bits.}.  If we restrict our analysis to findings with a severity greater than informational, SmartCheck detects no mutants of the contract, while Slither still reports that some mutants \emph{allow an arbitrary caller to cause the contract to destroy itself}.  Given that both tools, ignoring informational results, detect no problems with the original code, and only Slither detects any problems with the mutants, we can say that Slither performs better for this contract.

@@

A general objection to our approach is that mutants may differ substantially from ``real'' faults, in some way.  This is certainly true, in a sense \cite{GopinathMutants}, but for static analysis purposes we believe it does not matter.   The real risk is that some mutation operators align with patterns a particular tool identifies, biasing the evaluation in favor of that tool.  Such faults may be dis-proportionately present in mutants vs. real code.  However, we consider this unlikely.  The vast majority of applied mutation operations for all of our experiments were highly generic, and do not plausibly represent a pattern in which some tool might specialize.  Code deletions, the most common kind of mutation by far, leave no ``trace'' for a tool to match against, but only an omission, so cannot be subject to this concern.  Changing arithmetic and comparison operators and numeric constants (incrementing, decrementing, or changing to 0 or 1) account for most of the non-statement-deletion mutants, and it is difficult to imagine how any tool could unfairly identify these.

@@

\begin{itemize}[labelsep=3pt,leftmargin=12pt]
\item We propose a differential approach to comparing static analysis tools based on the insight that program mutants are easy-to-understand, likely-faulty, program changes.
\item We propose a definition of mutant killing for static analysis.
\item We introduce a simple scheme for prioritizing mutants that helps users understand and use the results of analysis.
\item We apply our method to an extensive, in-depth comparison of three Solidity smart contract analysis tools, and show how prioritization allowed us to easily identify (and build) three new detectors for the most effective of these tools.
\item We also provide results for popular Java and Python static analysis tools, further demonstrating our approach and showing strengths and weaknesses of these tools.
\end{itemize}

 While there are limitations to using differential mutation analysis to compare/improve static analysis tools, it scales to basing comparisons on many real software source files and very many ``faults,'' but still offers some of the advantages of having humans establish ground truth.

@@

\begin{enumerate}
\item Run each tool on the unmutated source code target(s), and determine the \emph{baseline}: the number of (non-informational/stylistic) findings produced.
\item Generate mutants of the source code and run each tool on each mutant.  Consider a mutant killed if the number of findings for the mutated code is greater than the number for the baseline, un-mutated code.
\item Compute, for each tool, the \emph{mutant ratio}:  the mutation score ($\frac{|\mathit{killed}|}{|\mathit{mutants}|}$) divided by (mean) baseline.  If it is zero, use a baseline equal to either one or the lowest non-zero baseline for any tool in the comparison set\footnote{This problem seldom arises in practice.}.
\item (Optional): Discard all mutants not killed by at least one tool and all mutants killed by all tools.  What remains allows \emph{differential} analysis.
Examine the remaining mutants in the difference in \emph{prioritized} order.
\end{enumerate}

@@

In order to apply FPF ranking to examining mutants, we implemented a
simple, somewhat \emph{ad hoc} distance metric and FPF ranker.  Our metric $d$ is
the sum of a set of measurements.  First, it adds a similarity
ratio based on Levenshtein distance \cite{lev} for (1) the \emph{changes} (Levenshtein edits) from
the original source code elements to
the two mutants,  (2) the two original source code elements changed (in
general, lines), and (3) the actual output mutant code.  These are
weighted with multipliers of 5.0, 0.1, and 0.1, respectively; the type
of change (mutation operator, roughly) dominates this part of the
distance, because it best describes ``what the mutant did''; however,
because many mutants will have the same change (e.g., changing {\tt +}
to {\tt -}, the other ratios also often matter.
Our metric also incorporates a measure of the distance in the source
code between the locations of two mutants.  If the mutants are to
different files, this adds 0.5; it also adds 0.025
times the number of source lines separating the two mutants if they
are in the same file, but caps the amount added at
0.25.

### 2021-04-20 relatedwork.tex
target ISSRE

Klinger et al. propose a different approach to differential testing of analysis tools \cite{klinger2019differentially}.  Their approach is in some ways similar to ours, in that it takes as input a set of seed programs, and compares results across new versions generated from seeds.  The primary differences are that their seed programs must be warning-free (which greatly limits the set of input programs available), and that the new versions are based on adding new assertions only.  We allow arbitrarily buggy seed programs, and can, due to the any-language nature of the mutation generator we use, operate even in new languages.  On the other hand, their approach can identify precision issues, while we offer no real help with false positives (in theory, you could apply their majority-vote method to mutants only a few tools flag, but mutants \emph{are} usually faults. in contrast to their introduction of checks that may be guaranteed to pass).  Most importantly, however, their approach \emph{only applies to tools that check assertions}, rather than the more common case of tools that identify bad code patterns (e.g. all tools we considered in this paper).

>>>>

Klinger et al. propose a different approach to differential testing of analysis tools \cite{klinger2019differentially}.  Their approach is in some ways similar to ours, in that it takes as input a set of seed programs, and compares results across new versions generated from seeds.  The primary differences are that their seed programs must be warning-free (which greatly limits the set of input programs available), and that the new versions are based on adding new assertions only.  We allow arbitrarily buggy seed programs, and can, due to the any-language nature of the mutation generator we use, operate even in new languages.  On the other hand, their approach can identify precision issues, while we offer no real help with false positives (in theory, you could apply their majority-vote method to mutants only a few tools flag, but mutants \emph{are} usually faults. in contrast to their introduction of checks that may be guaranteed to pass).  Most importantly, however, their approach \emph{only applies to tools that check assertions}, rather than the much more common case of tools that identify bad code patterns.
